---
title: "Introduction to the WRI Package"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
author: "Matthew Coleman, Lucy Liu and Alexander Petersen"
vignette: >
  %\VignetteIndexEntry{WRI-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(knitr)
library(statip)
library(ggplot2)
library(gridExtra)
library(ggpubr)
library(modeest)
```

The **W**asserstein **R**egression **I**nference (`WRI`) package is designed to perform statistical inference in the density regression, in which the response is a one-dimensional probability density and predictors are scalars. The package is based on the methods proposed in the paper, *Wasserstein F-tests and confidence bands for the Frechet regression of density response curves*. [Link to Paper on Arxiv](https://arxiv.org/abs/1910.13418).


Install and library `WRI` using:
```{r setup}
# install.packages('WRI')
library(WRI)
```

The dataset used in this description will be the `strokeCTdensity` dataset, which can be obtained through the **WRI** package. This dataset contains clinical, radiological scalar variables, and hematoma density curves for 393 stroke patients.

```{r}
data(strokeCTdensity)
?strokeCTdensity

predictor = strokeCTdensity$predictors
dSup = strokeCTdensity$densitySupport
densityCurves = strokeCTdensity$densityCurve
```

## Wasserstein Regression 

`wass_regress` is the estimation function which works similar to `lm`. To compute the fitted values, it requires a formula, response and predictor data. We give an explanation of other arguments below.  

*  `bandwidth`: the smoothing parameter in kernel smoothing applied on the density function, a value in (0, 1). Note that if the input density/quantile functions in `Ymat` are already smooth, setting `bandwidth = 0` will skip the smoothing.
*  `Ytype`: whether the response matrix contains `'quantile'` or `'density'` functions.
*  `Sup`: the common grid for density functions in `Ymat`.
*  `t`:  the common grid for quantile functions in `Ymat`. Note that even if the `Ymat` contains density functions, `t` can be specified properly to alleviate the effect of the sharp change in quantile density function. Please check the below illustration. 

### The effect of `t` grid vector

Since most derivation in WRI works in the space of quantile functions and its derivatives, the probability density function response needs to be converted into quantile function. However, the transformation will result in certain derivation between the original density function and $1/q(t)$, where $q(t) = Q'(t), t = F(x)$. Note that it is $q(t)$'s that are directly used in the WRI functions.

Below we set `t` as equally spaced grid vector and nonequally spaced vector, which is denser near the boundary to compare the resulting $1/q(t)$.  
```{r}
equal_t = den2Q_qd(densityCurves, dSup, bandwidth = 0, t = seq(0, 1, length.out = 120))
nonequal_t = den2Q_qd(densityCurves, dSup, bandwidth = 0, t = unique(c(seq(0, 0.05, 0.001), seq(0.05, 0.95, 0.05), seq(0.95, 1, 0.001))))
```

```{r, fig.align='center', fig.height=3, fig.width= 7.2, echo=F}
equal_t_1 = den2Q_qd(densityCurves, dSup, bandwidth = 0.1, t = seq(0, 1, length.out = 120))
nonequal_t_1 = den2Q_qd(densityCurves, dSup, bandwidth = 0.1, t = unique(c(seq(0, 0.05, 0.001), seq(0.05, 0.95, 0.05), seq(0.95, 1, 0.001))))
i = 1
df_t = data.frame(densityFun = c(densityCurves[i, ], equal_t$fobs_t[i, ], nonequal_t$fobs_t[i, ], 
                                 densityCurves[i, ], equal_t_1$fobs_t[i, ], nonequal_t_1$fobs_t[i, ]),
                  dSup = c(dSup,equal_t$Qobs_t[i, ], nonequal_t$Qobs_t[i, ],
                           dSup,equal_t_1$Qobs_t[i, ], nonequal_t_1$Qobs_t[i, ]),
                  density = c(rep('original f(x)', 101), rep('equal t: 1/q(t)', 120), rep('nonequal t: 1/q(t)', 119),
                              rep('original f(x)', 101), rep('equal t: 1/q(t)', 120), rep('nonequal t: 1/q(t)', 119)),
                  bandwith = rep(c('bandwidth = 0', 'bandwidth = 0.1'), each = 340))
ggplot(data = df_t, aes(x = dSup, y = densityFun, color = density)) +
  facet_grid(cols = vars(bandwith)) +
         geom_line() +
         geom_point(size = 0.3) +
         ggtitle('Comparison of the effect of equally and nonequally spaced t grid vector')

```

### Usage of `wass_regress` function

The density curves and predictor variables are input into `wass_regress` seperately. The Wasserstein regression can be performed with the following code:
```{r}
t = unique(c(seq(0, 0.05, 0.001), seq(0.05, 0.95, 0.05), seq(0.95, 1, 0.001)))
res = wass_regress(rightside_formula = ~., Xfit_df = predictor, Ymat = densityCurves, bandwidth = 0.01, Ytype = 'density', Sup = dSup, t = t)
```

The `wass_regress` function returns a WRI object. This object can be used with the other functions in this package to run hypothesis tests, calculate Wasserstein $R^2$, and compute confidence bands.


## Summary

The `summary` method for WRI objects combines the global F-test, Wassertstein $R^2$, and partial F-tests for individual effects into one easily-readable output.

```{r}
summary(res)
```

### Wasserstein Coefficient of Determination, $R^2$

The Wasserstein coefficient of determination, $R^2$ can be calculated with `wass_R2(res)`. The formula for the Wasserstein $R^2$ is as follows:

$$R^2=1-\frac{\sum^n_{i=1} d_W^2(f_{i}, \hat{f}_i)}{\sum^n_{i=1} d_W^2(f_{i}, \overline{f_{i}})},$$
Where $\overline{f_{i}}$ is the unconditional Wasserstein mean estimate and $\hat{f}_i$ is the conditional mean estimate.

This value represents the fraction of Wasserstein variability explained by the model, and can therefore be used to assess the goodness of fit for a model.

## Hypothesis Testing

### Global F-test

`WRI` also allows the computation of global F-tests through the `globalFtest` function. It provides four methods of computing the p-value, two (truncated and satterthwaite) through asymptotic analysis and two resampling techniques (permutation and bootstrap). Please note that the resampling methods can be slow. 

* Setting `permutation = TRUE` will also compute permutation p-value. The number of permutation samples can be controlled with the `numPermu` argument. 
* Setting `bootstrap = TRUE` will also compute bootstrap  p-value. The number of bootstrap samples can be controlled with the `numBoot` argument.

*Note on Degrees of Freedom* : The degrees of freedom are approximated by a chi-square distribution, so there is only 1 degree of freedom for our F-statistic. This is done because the F-statistic is asymptotically equivalent to a chi-squared distribution.

```{r}
globalFtable = globalFtest(res, alpha = 0.05, permutation = TRUE, numPermu = 200, bootstrap = TRUE, numBoot = 200)
kable(globalFtable$summary_df, digits = 3)
sprintf('The wasserstein F-statistic is %.3f on %.3f degrees of freedom', globalFtable$wasserstein.F_stat, globalFtable$chisq_df)
```

### Partial F-test

`partialFtest` can be used to test for individual effects or submodel fits. Using the stroke data as an example, we  test whether the clinical variables are significant for head CT hematoma densities when radiological variables are in the model.

```{r, eval=T}
# the reduced model only has four radiological variables
reduced_res = wass_regress(~ log_b_vol + b_shapInd + midline_shift + B_TimeCT, Xfit_df = predictor, Ymat = densityCurves, Ytype = 'density', Sup = dSup)
full_res = wass_regress(rightside_formula = ~., Xfit_df = predictor, Ymat = densityCurves, Ytype = 'density', Sup = dSup)

partialFtable = partialFtest(full_res, reduced_res, alpha = 0.05)
kable(partialFtable, digits = 3)
```

With p-value greater than 0.05, we are confident to conclude that when radiological variables are in the model, clinical variables are not significant for explaining the variance in head CT hematoma densities.

## Confidence Bands

### Usage of `confidenceBands` function
The `confidenceBands` function computes the intrinsic Wasserstein$-\infty$ bands and Wasserstein density bands. In the function, these refer to `quantile` band and `density` band respectively, which are controlled by `type` argument (options are 'quantile', 'density' or 'both'). By default, the function visualizes confidence bands for one object. But it allows to compute $m$ confidence bands simultaneously if a $m \times p$ dataframe `Xpred_df` is provided. All the results, including upper and lower bounds, predicted density function etc, are returned in a list.

```{r, fig.width=7.2, fig.height=2.5}
xpred = colMeans(predictor)
confidence_Band = confidenceBands(res, Xpred_df = xpred, type = 'both')
```

### Investigate the effect of hematoma volume

We set log(hematoma volume) equal to the first quartile (Q1) or third quartile (Q3) of the observed values, with all other predictors set at their mean (for continuous variables) or mode (for binary variables). Then compare the CT hematoma densities in these two cases.
```{r}
mean_Mode <- function(vec) {
  if (length(unique(vec)) < 3) {return(modeest::mlv(vec, method = "mfv"))}
  else {return(mean(vec))}
}
mean_mode_vec = apply(predictor, 2, mean_Mode)
predictorDF = rbind(mean_mode_vec, mean_mode_vec)
predictorDF[ , 1] = quantile(predictor$log_b_vol, probs = c(1/4, 3/4))
```

```{r, fig.width=7.2, fig.height=3}
res_cb = confidenceBands(res,predictorDF, level = 0.95, delta = 0.01, type = 'both', figure = F)
m = ncol(res_cb$quan_list$Q_lx)
na.mat = matrix(NA, nrow = 2, ncol = m - ncol(res_cb$den_list$f_lx))

cb_plot_df = with(res_cb, data.frame(
  fun = rep(c('quantile function', 'density function'), each = 2*m), 
  Q1Q3 = rep(rep(c('Q1', 'Q3'), each = m), 2),
  value_m = c(as.vector(t(quan_list$Qpred)), as.vector(t(cbind(cdf_list$fpred)))),
  value_u = c(as.vector(t(quan_list$Q_ux)), as.vector(t(cbind(den_list$f_ux, na.mat)))),
  value_l = c(as.vector(t(quan_list$Q_lx)), as.vector(t(cbind(den_list$f_lx, na.mat)))),
  support_full = c(rep(quan_list$t, 2), as.vector(t(cbind(cdf_list$Fsup)))),
  support_short = c(rep(quan_list$t, 2), as.vector(t(cbind(den_list$Qpred, na.mat))))
  ))
  
ggplot(data = cb_plot_df, aes(color = Q1Q3)) +
  geom_line(aes(x = support_full, y = value_m)) +
  geom_ribbon(aes(x = support_short, ymin = value_l, ymax = value_u, fill = Q1Q3), alpha = 0.25) +
  facet_wrap( ~ fun, scales = "free_y") +
  ylab('Confidence band') +
  xlab('Support')
```

